### 2015-06-05 chapters/ipa_meets_unicode.tex
major style revision

\section{Introducing the International Phonetic Alphabet (IPA)}
\label{introducing-the-international-phonetic-alphabet-ipa}

@@

\section{Pitfall: There is no single complete Unicode code block for IPA}
\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}

@@

\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> \item Latin-1
	Supplement (4 characters): <æ, ç, ð, ø\textgreater{} \item Latin
	Extended-A (3 characters): <ħ, ŋ, œ> \item Latin Extended-B (5
	characters): <ǀ, ǁ, ǂ, ǃ, > \item IPA Extensions (70 characters), e.g.:
	<ɐ, ɑ, ɔ> \item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥>
	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> \item
	Greek and Coptic (3 characters): <β, θ, χ> \item Phonetic Extensions (2
	characters): <, > \item Phonetic Extensions Supplement (3 characters): <
	, , > \item Superscripts and Subscripts (1 character): <ⁿ> \item Arrows
	(2 characters): \textless{}↑, ↓\textgreater{} \item Latin Extended-C (1
	character): <>
\end{itemize}

\section{Pitfall: There are many IPA homoglyphs in Unicode}
\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical, depending on the
font rendering). For example, a speaker of Russian should ideally not use the
\textsc{cyrillic small letter a} at code point U+0430 for IPA transcriptions,
but instead the \textsc{latin small letter a} at code point U+0061, although
visually they are mostly indistinguishable, and the Cyrillic character is more
easily typed on a Cyrillic keyboard. Another example we commonly encounter is
the use of \textsc{latin small letter g} at U+0067, instead of the sanctioned
Unicode Standard IPA character for the voiced velar stop \textsc{latin small
letter script g} at U+0261. One begins to question whether this issue is at all
apparent to the working linguist, or if they simply use the U+0067 because it is
easily keyboarded and thus saves time, whereas the latter must be cumbersomely
inserted as a special symbol in most software.\footnote{This issue was recently
addressed by the International Phonetic Association, which has taken the stance
that both the keyboard \textsc{latin small letter g} and the \textsc{latin small
letter script g} are valid input characters for the voiced velar plosive.
Unfortunately, this decision further introduces ambiguity for linguists trying
to adhere to a strict Unicode Standard IPA encoding.}

Furthermore, on the one hand even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point U+0259 and <ǝ>
\textsc{latin small letter turned e} at U+01DD. On the other hand, non-linguists
are unlikely to distinguish any semantic difference between an open back
unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at U+0251, and the
open front unrounded vowel <a>, \textsc{latin small letter a} at U+0061. But
even among linguists this distinction leads to problems. For example, as pointed
out by \citet{Mielke2009}, there is a problem stemming from the fact that about
75\% of languages are reported to have a five-vowel system
\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
precise audio recording and measurements of formants, so this may lead one to
ask if the many characters that are used in phonological description reflects a
\textsc{transcriptional bias}. The common use of in transcriptions could be in
part due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'', at:
\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
Making things even more problematic, there is an old typographic tradition that
the double-story uses a single-story in italics. This leads to the unfortunate
effect that in most well-designed fonts the italics of and use the same glyph.
If this distinction has to be kept upright in italics, the only solution we can
currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
glyphs) instead of real italics (i.e.~special italics glyphs designed by a
typographer).\footnote{For example, the widely used IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect. However, note that
when the text is transferred to another font (i.e.~one that includes real
italics) and/or to another software application (like Apple Pages, which does
not perform slanting), then this visual appearance will be lost. In this case we
are thus still in the pre-Unicode situation in which the choice of font and
rendering software actually matters. The ideal solution from a linguistic point
of view would be the introduction of a new IPA code point for a different kind
of which explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a>-like characters in Unicode (which
already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
clear situation in which the Unicode Consortium is not just thinking about
linguists, but has a more wide-ranging practical view to consider.}

@@

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the
	Unicode Standard email list. An English keyboard inputs
	\textless{}`\textgreater{} APOSTROPHE at U+0027, although the
	\textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
	\textsc{right single quotation mark} at U+2019. Yet the glottal
	stop/glottalization/ejective marker is another completely different
	character, the \textsc{modifier letter apostrophe} at U+02BC, but
	unfortunately looks mostly highly similar to U+2019.  \item There is
	ambiguous encoding of IPA segments within the Unicode Standard. An
	example is the U+02C1 \textsc{modifier letter reversed glottal stop} vs
	the U+02E4 \textsc{modifier letter small reversed glottal stop} . Both
	are denoted in the Unicode Standard as the \textsc{pharyngealized
	diacritic} and both appear in various resources representing phonetic
	data online. This is thus an example for which the Unicode Standard does
	not solve the linguistic standardization problem.  \item There is at
	least one case in which the character name assigned by the Unicode
	Consortium does not match the IPA's description: in the Unicode Standard
	at U+01C3 is labeled \textsc{latin letter retroflex click}, but in IPA
	is an alveolar or postalveolar click (not retroflex). This naming is
	probably best seen as a simple error in the Unicode Standard. Note that
	most linguists simply seem to use or <ou>) while ligatures are single
	characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at
	U+02A7). Ligatures arose in the context of printing easier-to-read
	texts, and are included in the Unicode Standard for reasons of legacy
	encoding. However, their usage is discouraged by the Unicode core
	specification. Specifically related to IPA, various phonetic
	combinations of characters (typically affricates) are available as
	single code-points in the Unicode Standard, but are designated as
	\textsc{ligatures} or \textsc{digraphs} (confusingly both names appear
	interchangeably). Such glyphs might be used by software to produce a
	pleasing display, but they should not be hard-coded into the text
	itself. In the context of IPA, characters like the following ligatures
	should thus \emph{not} be used. Instead a combination of two characters
	is preferred: \item <ʣ> \textsc{latin small letter dz digraph} at U+02A3
	(use <d> + <z> instead) \item <ʧ> \textsc{latin small letter tesh
	digraph} at U+02A7 (use <t> + <ʃ> instead) \item <ʩ> \textsc{latin small
	letter feng digraph} at U+02A9 (use <f> + <ŋ> instead)
\end{itemize}

@@

\begin{itemize}
	\item <ɮ> \textsc{latin small letter lezh} at U+026E \item <œ>
	\textsc{latin small ligature oe} at U+0153 \item <ɶ> \textsc{latin
	letter small capital oe} at U+0276 \item <æ> \textsc{latin small letter
	ae} at U+00E6
\end{itemize}

\section{Pitfall: The notion of diacritic differs between IPA and Unicode}
\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}

@@

\begin{itemize}
	\item Length marks, namely <ː> \textsc{modifier letter triangular colon}
	at U+02D0 and <ˑ> \textsc{modifier letter half triangular colon} at
	U+02D1.  \item Tone letters, like <˥> \textsc{modifier letter extra-high
	tone bar} at U+02E5, and others like this.  \item Superscript letters,
	like <ʰ> \textsc{modifier letter small h} at U+02B0 or <ˤ>
	\textsc{modifier letter small reversed glottal stop} at U+02E4, and
	others like this.  \item <˞> \textsc{modifier letter rhotic hook} at
	U+02DE.
\end{itemize}

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
U+02B0 the Unicode Consortium's decision to treat it as a separate character is
also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there is a mechanism in
Unicode to force separate characters to be combined (namely by using the
\textsc{zero width joiner} at U+200D), but this seems to be a rather
impractical, and probably not enforceable solution to us.

\section{Pitfall: There is no unique diacritic ordering in IPA, nor in Unicode}
\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}

@@

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization.
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at U+00E1) will be split into its parts. This might sound
preferable for a linguistic analysis, as the different diacritics are separated
from the base characters. However, note that most attached elements like strokes
(e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>)
will not be decomposed, but strangely enough a cedilla (like in <ç>) will be
decomposed. In general, Unicode decomposition does not behave like a feature
decomposition as expected from a linguistic perspective. It is thus important to
consider Unicode decomposition only as a technical procedure, and not assume
that it is linguistically sensible.

>>>>

\section{The International Phonetic Alphabet}
\label{the-international-phonetic-alphabet}

@@

\section{Pitfall: No complete IPA code block}
\label{pitfall-no-complete-ipa-block}

@@

\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
	\item Greek and Coptic (3 characters): <β, θ, χ> 
	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
	\item Superscripts and Subscripts (1 character): <ⁿ> 
	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
	\item Latin Extended-C (1 character): <ⱱ> 
\end{itemize}

\section{Pitfall: Many IPA homoglyphs}
\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical, depending on the
font rendering). For example, a speaker of Russian should ideally not use the
\textsc{cyrillic small letter a} at code point \uni{0430} for IPA
transcriptions, but instead the \textsc{latin small letter a} at code point
\uni{0061}, although visually they are mostly indistinguishable, and the
Cyrillic character is more easily typed on a Cyrillic keyboard. Another example
we commonly encounter is the use of \textsc{latin small letter g} at \uni{0067},
instead of the sanctioned Unicode Standard IPA character for the voiced velar
stop \textsc{latin small letter script g} at \uni{0261}. One begins to question
whether this issue is at all apparent to the working linguist, or if they simply
use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
the latter must be cumbersomely inserted as a special symbol in most
software.\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}

Furthermore, on the one hand even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. On the other hand,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a>, \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a \textsc{transcriptional bias}. The common use of in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'', at:
\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
Making things even more problematic, there is an old typographic tradition that
the double-story uses a single-story in italics. This leads to the unfortunate
effect that in most well-designed fonts the italics of and use the same glyph.
If this distinction has to be kept upright in italics, the only solution we can
currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
glyphs) instead of real italics (i.e.~special italics glyphs designed by a
typographer).\footnote{For example, the widely used IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect. However, note that
when the text is transferred to another font (i.e.~one that includes real
italics) and/or to another software application (like Apple Pages, which does
not perform slanting), then this visual appearance will be lost. In this case we
are thus still in the pre-Unicode situation in which the choice of font and
rendering software actually matters. The ideal solution from a linguistic point
of view would be the introduction of a new IPA code point for a different kind
of which explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a>-like characters in Unicode (which
already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
clear situation in which the Unicode Consortium is not just thinking about
linguists, but has a more wide-ranging practical view to consider.}

@@

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{} \textsc{right single quotation mark} at \uni{2019}. Yet the glottal stop/glottalization/ejective marker is another completely different character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the \uni{02C1} \textsc{modifier letter reversed glottal stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal stop} . Both are denoted in the Unicode Standard as the \textsc{pharyngealized diacritic} and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click}, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs} (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} (use <t> + <ʃ> instead) 
	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9} (use <f> + <ŋ> instead) 
\end{itemize}

@@

\begin{itemize}
	\item <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
	\item <œ> \textsc{latin small ligature oe} at \uni{0153} 
	\item <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
	\item <æ> \textsc{latin small letter ae} at \uni{00E6} 
\end{itemize}

\section{Pitfall: No decomposition of strikethrough}
\label{pitfall-no-decomposition-of=strikethrough}

\section{Pitfall: Different notions of diacritics}
\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}

@@

\begin{itemize}
	\item Length marks, namely <ː> \textsc{modifier letter triangular colon} at \uni{02D0} and <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}. 
	\item Tone letters, like <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}, and others like this. 
	\item Superscript letters, like <ʰ> \textsc{modifier letter small h} at \uni{02B0} or <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}, and others like this. 
	\item <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}. 
\end{itemize}

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there is a mechanism in
Unicode to force separate characters to be combined (namely by using the
\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
impractical, and probably not enforceable solution to us.

\section{Pitfall: No unique diacritic ordering}
\label{pitfall-no-unique-diacritic-ordering}

@@

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization.
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed. In general, Unicode decomposition does not behave like
a feature decomposition as expected from a linguistic perspective. It is thus
important to consider Unicode decomposition only as a technical procedure, and
not assume that it is linguistically sensible.

### 2015-06-05 chapters/orthography_profiles.tex
major style revision

Given these goals, Unicode locale descriptions (see Section \ref{terminology}) might seem like the ideal orthography profiles. However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description. This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository. The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).

The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system. Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies. Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved. In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g. between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/). In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted. To extract data for analysis, a source-by-source approach is required before an orthography profile can be created. For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.

@@

As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :

@@

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space. Unicode tokenization is required because sequences of code points can differ in their visual and logical orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs, computationally they are different. Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison. In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard. The combining character sequence <õ̰> is normalized and visually grouped together. Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with . This is because it belongs to Spacing Modifier Letters category in the Unicode Standard. These characters are underspecified for the direction in which they modify a host character. For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character). Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile). For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme . The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.

\section{Informal description of orthography profiles}
\label{informal-description-of-orthography-profiles}

@@

In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected. In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section \ref{use-cases} on various exemplary use-cases). The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile. Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.

\section{Formal specification of orthography profiles}
\label{formal-specification-of-orthography-profiles}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section \ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography. 
	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}. A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header. For all columns we assume the name in the header of the CSV file to be crucial. The actual ordering of the columns is unimportant. 
	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at U+0023). Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}. Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.} 
	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format. }Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile. 
\end{enumerate}

>>>>

Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
might seem like the ideal orthography profiles. However, there are various
practical obstacles preventing the use of such locale descriptions in the daily
linguistic practice, namely: (i) the XML-structure is too verbose to easily and
quickly produce or correct manually, (ii) locale descriptions are designed for a
wide scope on information (like date formats or names of weekdays) most of which
is not applicable for documenting writing systems, and (iii) most crucially,
even if someone made the effort to produce a technically correct locale
description for a specific source at hand, then it is nigh impossible to deploy
the description. This is because a locale description has to be submitted to and
accepted by the Unicode Common Locale Data Repository. The repository is
(rightly so) not interested in descriptions that only apply to a limited set of
sources (e.g.~only a single specific dictionary).

The major challenge then is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
a nasalized vowel /ṽ/). In our experience, data from each source must be
individually tokenized into graphemes so that its orthographic structure is
identified and its contents can be extracted. To extract data for analysis, a
source-by-source approach is required before an orthography profile can be
created. For example, almost each available lexicon on the world's languages is
idiosyncratic in its orthography and thus requires lexicon-specific approaches
to identify graphemes in the writing system and to map graphemes to phonemes, if
desired.

@@

As an example of graphemic tokenization, note the three different levels of
technological and linguistic elements that interact in the hypothetical lexical
form <tsʰṍ̰shi>:

@@

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> +
<̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs,
computationally they are different. Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison. In (2), the
Unicode code points have been logically normalized and visually organized into
grapheme clusters, as specified by the Unicode Standard. The combining character
sequence <õ̰> is normalized and visually grouped together. Note that, the
MODIFIER LETTER SMALL H at \uni{02B0}, is not grouped with. This is because it
belongs to Spacing Modifier Letters category in the Unicode Standard. These
characters are underspecified for the direction in which they modify a host
character. For example, can indicate either pre- or post-aspiration (whereas the
nasalization or creaky diacritic is defined in the Unicode Standard to apply to
a specified base character). Finally, to arrive at the graphemic tokenization in
(3), tailored grapheme clusters are needed (as for example specified in an
orthography profile). For example, this orthography profile would specify that
the sequence of characters, and form a single grapheme, and that and form a
grapheme. The orthography profile could also specify orthographic rules,
e.g.~when tokenization graphemes, in say English, the in the forms and should be
treated as distinct sequences depending on their contexts.

\section{Informal description}
\label{informal-description-of-orthography-profiles}

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Section~\ref{use-cases} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

\section{Formal specification}
\label{formal-specification-of-orthography-profiles}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section~\ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography. 
	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}. A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header. For all columns we assume the name in the header of the CSV file to be crucial. The actual ordering of the columns is unimportant. 
	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at \uni{0023}). Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}. Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.} 
	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format. }Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile. 
\end{enumerate}
